---
title: "Development of a dynamic model for the emergence of Lassa fever in West Africa" 
subtitle:
  - "Draft Thesis November 2022"
author:
  - "David Simons"
date:
  - "November 2022"
site: bookdown::bookdown_site
output: 
  bookdown::pdf_document2:
    includes:
      in_header: "preamble.tex"
bibliography: ["references/references.bib"]
csl: "harvard.csl"
lof: true
lot: true
space_between_paragraphs: true
---

```{r knitr-options, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, dpi = 320, 
                      fig.width = 8, fig.height = 8,
                      out.width = "80%", fig.align = 'center')
```

```{r include_packages, include = FALSE}
if (!require(devtools))
  install.packages("devtools", repos = "http://cran.rstudio.com")
if (!require(thesisdown))
  devtools::install_github("ismayc/thesisdown")
library(thesisdown)
if (!require(pacman))
  install.packages("pacman", repos = "http://cran.rstudio.com")
pkgs =
  c("ggspatial",
    "googledrive",
    "here",
    "mapdata",
    "maps",
    "maptools",
    "sf",
    "tidyterra",
    "tidyverse")

pacman::p_load(pkgs, character.only = T)
```


A dissertation submitted in partial fulfilment of the requirements for the degree of **Doctor of Philosophy**

## Declaration {-}

I certify that:   

+ The thesis being submitted for examination is my own account of my own research;
+ My research has been conducted ethically;
+ Where I have drawn on the work, ideas and results of others this has been appropriately cited in the thesis;
+ Where any collaboration has taken place with other researchers, I have clearly stated in the thesis my own personal contribution;
+ The entirety of the work described in the thesis has been undertaken subsequent to my registration for the higher degree for which I am submitting for examination;
+ The thesis submitted is within the required word limit as specified by the RVC.

## Abstract {-}

```{r abstract, child = "00-abstract.Rmd"}
```

## Impact statement {-}

```{r impact, child = "00-impactstatement.Rmd"}
```

## List of Acronyms {-}

```{r acronyms, child = "00-acronyms.Rmd"}
```

## Definitions used {-}

```{r definitions, child = "00-definitions.Rmd"}
```

## Acknowledgements {-}

```{r acknowledgements, child = "00-acknowledgements.Rmd"}
```

## Chapter overview and collaborators {-}

```{r collaborators, child = "00-chapteroverview.Rmd"}
```

## Thesis output {-}

```{r output, child = "00-output.Rmd"}
```