\begin{titlepage}

\begin{centering}

\huge

\doublespacing

\vspace*{2cm}

{\bf Investigating rodent ecology to understand the hazard of Lassa fever spillover in Sierra Leone}

\vspace{1cm}

\large

{\em David Simons}

\vspace{2cm}

A dissertation submitted in partial fulfilment \\of the requirements for the degree of \\{\bf Doctor of Philosophy}

\vspace{1.5cm}

The Royal Veterinary College, University of London

\normalsize
July 2023

\vspace{1.5cm}

{\em The copyright in this thesis is owned by the author. Any quotation from the thesis or use of any of the information contained in it must acknowledge this thesis as the source of the quotation or information.}

\end{centering}

\end{titlepage}

\newpage

\Large

Declaration

\normalsize

I certify that:  

\begin{itemize}
  \item The thesis being submitted for examination is my own account of my own research;  
  \item My research has been conducted ethically;  
  \item Where I have drawn on the work, ideas and results of others this has been appropriately cited in the thesis;  
  \item Where any collaboration has taken place with other researchers, I have clearly stated in the thesis my own personal contribution;  
  \item The entirety of the work described in the thesis has been undertaken subsequent to my registration for the higher degree for which I am submitting for examination;  
  \item The thesis submitted is within the required word limit as specified by the RVC.  
\end{itemize}

\newpage

\Large

Acknowledgements

\normalsize

First of all, I would like to thank my fantastic supervisors: Richard Kock, Deborah Watson-Jones and Kate Jones. I have learnt so much from your expertise, detailed inputs and scientific integrity, and I am deeply grateful for your support and mentorship. I am particularly thankful to Richard for the numerous opportunities you have provided and for bringing me into the PANDORA-ID-NET consortium. I would similarly like to thank Deborah for your support in conducting research in West Africa and assistance integrating a medical perspective into the One Health approach taken throughout my research. I am also thankful to  Kate for your methodological expertise, advancing my understanding of ecology and helping me to develop as a scientist. It has been an absolute privilege to work with such inspiring leaders in the fields of Veterinary Epidemiology, Clinical Epidemiology and Ecology. I would also like to thank the London Interdisciplinary Biosciences Consortium, Doctoral Training Programme for funding my PhD and the PANDORA-ID-NET programme for support conducting my research in Sierra Leone.
\par
Thank you also to the communities in Sierra Leone within which I worked. Without the support of the communities of Baiama, Lalehun, Lambayama and Seilama and the wider Kenema Chiefdom this work would not have been possible, and I am extremely grateful for how welcome you made me and the field team in the pursuit of this work. Thank you to my collaborators in Sierra Leone: Dianah Sondufu, Joyce Lamin, Michael Dawson, "Bobby", Joseph Lahai, "Coach", James Koninga, Momoh Jimmy and Rashid Ansumana. I am deeply grateful for your intellectual and practical inputs alongside your humour when things did not go exactly to plan.
\par
Thank you to everyone in the PANDORA-ID-NET consortium, Kate Jones lab at UCL and the Elisabeth Fichet-Calvet lab at BNITM, I have thoroughly enjoyed our discussions which have helped to shape this thesis. I feel lucky to have been part of such dedicated and supportive research groups. In particular heartfelt thanks to Umaru Bangura and Rory Gibb for your invaluable methodological inputs.
\par
Thank you to my family and friends for your love and support through my never-ending journey in education. I would like to express my gratitude to my partner in science and life Olga for her constant encouragement. Without you, this PhD would not have seemed possible and would likely have remained incoherent musings on zoonotic diseases.

\newpage